Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-04 doc/invt.tex
add sinc.gp

Assuming the system is initially equilibrated
under a constant updating magnitude, $\alpha_0$,
we may use Eq. \eqref{eq:error_res1} with
$$
q(T)
=
\int \frac{ dt } { t + t_0 }
=
\ln \frac{ T + t_0 } { t_0 }.
$$
%
Thus, from Eqs. \eqref{eq:error_split}
and \eqref{eq:error_asym_sinc}, we have
%
\begin{equation}
  E
  =
  E_R + E_A
  =
  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
        {          ( T + t_0 )^2          }
  \sum_{ k = 2 }^K \Gamma_k
  .
\notag
%\label{eq:error_sinc0}
\end{equation}
%
The optimal error is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{        1         }
       { T + 2 / \alpha_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This means that at the beginning of the schedule,
the updating magnitude is
$\alpha(0^+) = \alpha_0 / 2$,
half of the previous equilibrium value.
%
Interestingly, this factor of $1/2$
happens to coincide with the suggested
reduction factor of updating magnitude
in the original WL algorithm\cite{wang2001, wang2001pre}
during stage transitions.

@@

Real simulations, however,
can be relatively short
and need not enter
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(T) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ T + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (T + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { T + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (T + t_0)    }
,
$$
is the equilibrium, or saturated, error of mode $k$
at the terminal updating magnitude
of $\alpha(T) = 1/[\lambda (T + t_0)]$.

\note{Derivation of Eq. \eqref{eq:error_invt}:
$$
\begin{aligned}
\Err_R
&=
\sum_{ k = 2 }^n
  \Err^f_k \left( \frac{   t_0   }
                       { T + t_0 }
          \right)^{ 2 \nu_k - 1 }
\\
\Err_A
&=
\sum_{ k = 2 }^n
  \frac{ 2 \, \Err^f_k \, \nu_k }
       {    2 \, \nu_k - 1      }
  \left[
    1 -
    \left(
      \tfrac{   t_0   }
            { T + t_0 }
      \right)^{2 \, \nu_k - 1}
  \right]
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left[
    1 -
    \left(
      \tfrac{   t_0   }
            { T + t_0 }
      \right)^{2 \, \nu_k - 1}
  \right]
  +
  \frac{   \Err^f_k     }
       { 2 \, \nu_k - 1 }
  \left[
    1 -
    \left(
      \tfrac{   t_0   }
           { T + t_0 }
      \right)^{2 \, \nu_k - 1}
  \right].
\end{aligned}
$$
Adding the two yields Eq. \eqref{eq:error_invt}.
}

@@

Clearly, the final error is always greater than
the saturated equilibrium error
at the terminal $\alpha(T)$:
$$
E \ge E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
$$
because the second term in the braces
of Eq. \eqref{eq:error_invt}
is nonnegative no matter the sign of $2 \, \nu_k - 1$
(in fact, it is a decreasing function
that vanishes only at $\nu_k \to +\infty$).
%
For a set of nonidentical $\lambda_k$'s,
the equality is reached only at $T = 0$.

@@

Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{    \Gamma_k    }
       { 2 \, (T + t_0) }
  G\left( \nu_k, \frac{ t_0 } { T + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\nu_k
\,
\left(
  1
  +
  \frac{ 1 - r^{2 \, \nu_k - 1} }
       {   2 \, \nu_k - 1       }
\right).
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (T + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$,
and the other shallower one around $\nu_k = 1$.
%
For a short simulation, or a large $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$, or increasing simulation length,
under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error
is always achieved at the other minimum,
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.

>>>>

Assuming the system is initially equilibrated
under a constant updating magnitude, $\alpha_0$,
we may use Eq. \eqref{eq:error_res1} with
$$
q(T)
=
\int \frac{ dt } { t + t_0 }
=
\ln \frac{ T + t_0 } { t_0 }.
$$
%
Thus, from Eqs. \eqref{eq:error_split}
and \eqref{eq:error_asym_sinc}, we have
%
\begin{equation}
  E
  =
  E_R + E_A
  =
  \frac { \frac 1 2 \alpha_0 \, t_0^2 + T }
        {          ( T + t_0 )^2          }
  \sum_{ k = 2 }^K \Gamma_k
  .
\notag
%\label{eq:error_sinc0}
\end{equation}
%
The optimal error is reached at
%
\begin{equation}
  t_0
  =
  \frac{     2    }
       { \alpha_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
and at this point
%
\begin{equation}
  E
  =
  \frac{        1         }
       { T + 2 / \alpha_0 }
  \sum_{ k = 2 }^K
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This means that at the beginning of the schedule,
the updating magnitude is
$\alpha(0^+) = \alpha_0 / 2$,
half of the previous equilibrium value.
%
Interestingly, this factor of $1/2$
happens to coincide with the suggested
reduction factor of updating magnitude
in the original WL algorithm\cite{wang2001, wang2001pre}
at stage transitions.

@@

Real simulations, however,
can be relatively short
and need not enter
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(T) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ T + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { T + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \frac
  {
    \Gamma_k \, \nu_k \,
    \left[
      \nu_k
      -
      (\nu_k - 1)
      \left(
        \frac{ t_0 } { T + t_0 }
      \right)^{ 2 \, \nu_k - 1 }
    \right]
  }
  {
    (T + t_0) \, (2 \, \nu_k - 1)
  }
  .
\label{eq:error_invt}
\end{align}
%
%where
%$$
%\Err^f_k
%\equiv
%\left\langle
%  y_k^2
%\right\rangle_{ \alpha(t) }
%=
%\frac{  \Gamma_k \, \nu_k   }
%     {    2 \, (T + t_0)    }
%,
%$$
%is the equilibrium, or saturated, error of mode $k$
%at the terminal updating magnitude
%of $\alpha(T) = 1/[\lambda (T + t_0)]$.

%\note{Derivation of Eq. \eqref{eq:error_invt}:
%$$
%\begin{aligned}
%\Err_R
%&=
%\sum_{ k = 2 }^n
%  \Err^f_k \left( \frac{   t_0   }
%                       { T + t_0 }
%          \right)^{ 2 \nu_k - 1 }
%\\
%\Err_A
%&=
%\sum_{ k = 2 }^n
%  \frac{ 2 \, \Err^f_k \, \nu_k }
%       {    2 \, \nu_k - 1      }
%  \left[
%    1 -
%    \left(
%      \tfrac{   t_0   }
%            { T + t_0 }
%      \right)^{2 \, \nu_k - 1}
%  \right]
%\\
%&=
%\sum_{ k = 2 }^n
%  \Err^f_k \,
%  \left[
%    1 -
%    \left(
%      \tfrac{   t_0   }
%            { T + t_0 }
%      \right)^{2 \, \nu_k - 1}
%  \right]
%  +
%  \frac{   \Err^f_k     }
%       { 2 \, \nu_k - 1 }
%  \left[
%    1 -
%    \left(
%      \tfrac{   t_0   }
%           { T + t_0 }
%      \right)^{2 \, \nu_k - 1}
%  \right].
%\end{aligned}
%$$
%Adding the two yields Eq. \eqref{eq:error_invt}.
%}

@@

%Clearly, the final error is always greater than
%the saturated equilibrium error
%at the terminal $\alpha(T)$:
%$$
%E \ge E_\mathrm{sat} \equiv \sum_{k = 2}^n E^f_k,
%$$
%because the second term in the braces
%of Eq. \eqref{eq:error_invt}
%is nonnegative no matter the sign of $2 \, \nu_k - 1$
%(in fact, it is a decreasing function
%that vanishes only at $\nu_k \to +\infty$).
%%
%For a set of nonidentical $\lambda_k$'s,
%the equality is reached only at $T = 0$.

@@

Let us first rewrite Eq. \eqref{eq:error_invt} as
%
$$
\Err
=
\sum_{ k = 2 }^n
  \frac{ \Gamma_k  }
       { (T + t_0) }
  G\left( \nu_k, \frac{ t_0 } { T + t_0} \right),
$$
%
where the $k$th component is proportional to
%
$$
G\left( \nu_k, r \right)
=
\frac { \nu_k }
      { 2 \, \nu_k - 1 }
\,
\left[
  \nu_k
  -
  (\nu_k - 1) \,
  r^{2 \, \nu_k - 1}
\right].
$$
%
As shown in Fig. \ref{fig:err_component},
the function $G(\nu_k, r)$ depends non-trivially on $\nu_k$.
%
For a fixed and sufficiently small $r = t_0 / (T + t_0)$,
$G(\nu_k, r)$ has two local minima along $\nu_k$,
one at $\nu_k = 0$,
and the other shallower one around $\nu_k = 1$.
%
For a short simulation, or a large $r$,
the minimal error may be achieved at $\nu_k \approx 0$,
corresponding to a large value of $\lambda$
($\lambda \gg \lambda_k$).
%
However, this minimum is unstable,
for $G(\nu_k, r)$ increases rapidly
with decreasing $r$, or increasing simulation length,
under a fixed $\nu_k$.
%
In other words,
asymptotically, the minimal error
is always achieved at the other minimum,
$\nu_k = 1$ or $\lambda = \lambda_k$.
%
In this case,
$G(\nu_k, r) \to G(1, 0) = 2$
and the total error is roughly twice $E_\mathrm{sat}$.
